Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-11 paper/paper.tex
Fixed issues in op-semantics after discussion with Sanyal Sir
Updated op-semantics and minefield semantics
reorg of fig
appendix is back

Functional programs extensively use dynamically allocated memory.  The
allocation  is either  explicit (for  example, using  constructors) or
implicit  (during creation  of  closures).  Programs  written in  lazy
functional languages put additional  demands on memory as they require
closures to  be carried  from the  point of creation  to the  point of
evaluation.

@@

\begin{figure*}[t!]
\begin{center}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
\\
\hline
          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
\rightsquigarrow
$  $\rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e$      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
\rightsquigarrow
\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
\sc{funcall} \\
\hline
$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
e)\!:\!S, H,
(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
\end{figure*}

@@

%------------------------------------------------------------%
\subsubsection{From CFGs to GC-ready DFA}
\label{sec:NFA-approx}
 We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
 approximate a CFG to a {\em strongly regular\/} grammar.  For
 example, the grammar
 fragment  for  the  non-terminal  $\var{\Df{\length}{1}}$  after  the
 Mohri-Nederhof transformation is:
 \begin{eqnarray*}
   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
   \acdr\var{\Df{\length}{1}}
   \mid \clazy\var{\Df{\length}{1}}\\
   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
   \mid \epsilon
 \end{eqnarray*}

The strongly regular grammar is converted  into a set of NFAs, one for
each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
pairs  of consecutive  edges labelled  \bcar\acar\ or  \bcdr\acdr\ and
constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
simplification is continued till a fixed point is reached, after which
the edges labelled \bcar \ and  \bcdr are deleted.  The details of the
algorithm, its correctness and termination proofs are given by Karkare
et.  al.~\cite{karkare07liveness,asati14lgc}.  The resulting automaton
has  edges labelled  with  \acar,  \acdr\ and  \clazy\  only. In  this
automaton, for every edge labelled \clazy, we check if the source node
of the edge has a valid path to  a final final state.  If yes, we mark
the mark  the source node as  final. Finally, we remove  all the edges
labelled  \clazy\  and  convert  the automaton  into  a  deterministic
automaton.    This   effectively  implements   the   $\hookrightarrow$
simplification rules for  \bcar, \bcdr, and \clazy\  to obtain forward
paths. While  checking for liveness  during GC, a  forward access-path
only if it can reach a final state.

@@

Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
from the grammar for \var{\Df{\length}{1}}.  The final DFA is shown in
Figure~\ref{fig:example-automata}(b).  This expectedly says that given
a  demand $\epsilon$  on  \length,  the liveness  of  its argument  is
$\acdr^{*}$  (the  spine  of   the  list  is  traversed).   Similarly,
Figure~\ref{fig:example-automata}(c)     shows     the     NFA     for
\var{\Lanv{9}{\pa}}.          The         DFA         shown         in
Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
paths.   This  reflects  the  lazy  nature  of  our  language.   Since
\length\ does not  evaluate the member elements of  the argument list,
the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
liveness based GC is triggered beyond $\pi_9$.

@@

\mycomment{
\pagebreak
\pagebreak
\section{Appendix}

@@

To set up the minefield semantics, we follow these steps:
\begin{enumerate}
\item We enrich  the abstract machine state $\rho,$ $S,$  $ H,$ $e$ to
  $\rho,$ $S,$ $H,$  $e,$ $\sigma,$ $e_f$.  $\sigma$ is  the demand on
  the expression and $e_f$ is the  body of the function.  The body and
  the demand are both needed to calculate the liveness environment for
  expressions.  The evaluation of  an expression may make variables in
  the scope live because of closures that are created outside it.  The
  information  in a  suspended  evaluation context  is also  similarly
  augmented with the demand on it and the function body of which the
  suspended evaluation is a part.  Thus a stacked entry now takes the
  form $(\rho, x, e, \sigma, e_f)$. A modification of the small step
  semantics which carries the extra information is shown in Figure[].
\item Given the current evaluation context $\rho, S, H, e, \sigma,
  e_f$ and $LF$, we can construct a liveness environment as follows:
  \begin{enumerate}
  \item If $e$ is an expression, then:  
  $$ \Lv = \bigcup_{\psi \in \mathcal{P}(e)} \mathcal{M}(\psi),
    \text{where }
  \mathcal{M} = \mathcal{L}(e_f, \sigma, \Lfonly)$$
\item If $e$ is an application, then:  
  $$ \Lv = ref(e, \sigma, \Lfonly)$$
  \end{enumerate}
We can define in the same way the liveness environment for each of the
suspended  evaluation contexts in  $S$  giving  a  stack of  liveness
environments that we shall denote $\mathsf{SL}$.
\item $GC$ models a liveness based garbage collection:
$GC(\rho,$ $ S,$ $ H,$ $ L,$ $ SL) = (\rho', S', H')$ 
  \begin{enumerate}
  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
    \notin \Lv$.
\item For each stack entry $(\rho,\_,\_,\_,\_)$ in $S$ with $\Lv'$ as
  the corresponding liveness environment in $\mathsf{SL}$,  and for each $x
  \in domain(\rho)$,  $\rho'(x)=\bot$
  iff $x.\epsilon \notin \Lv'$.
\item For each location $\ell$, $H'(\ell) = \bot$, iff  there is no
  $x$ in either the current environment or one of the stacked
  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
  \ell$  and $x.\alpha \in \Lv$, .
  \end{enumerate}
\end{enumerate}

\begin{figure*}[h!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\ 
\hline
\hline
          &\makecell{ $\rho, (\rho', x, e, \cred{\sigma', e_f'})\!:\!S,
  H, \kappa, \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) :=
    \kappa], e, \cred{\sigma', e_f'}$ }   &  \sc{const}
\\
\hline
          & \makecell{$\rho, (\rho', z, e, \cred{\sigma',
    e_f'})\!:\!S, H, (\CONS~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e,
  \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \cred{\sigma, e_f}$ \\ $
  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e, \cred{\sigma', e_f'}$}      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
  H, (\CAR~x), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
  (\CAR~x), \cred{\sigma, e_f})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma , \_ }$}      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \makecell{$\rho, (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H,
  (+~x~y), \cred{\sigma, e_f}$ \\  $\rightsquigarrow \rho', S, H[\rho(z)
    \mapsto H(\rho(x)) + H(\rho(y))], e, \cred{\sigma', e_f'}$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
  H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_}$}      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
  S, H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, y,
  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_ }$}      &
\sc{prim-2-clo} \\
\hline
{$\mathit{g}~\mbox{defined as}$
$~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
  (g~\myvec{x}), \cred{\sigma, e_f}$ \\ $\rightsquigarrow [\myvec{y} \mapsto
    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma, e_g}$}      &
\sc{funcall} \\
\hline
$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma, e_f}$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e, \cred{\sigma, e_f}$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma, e_f}$} & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma, e_f}$ \\   $\rightsquigarrow
\rho, S, H,  e_2, \cred{\sigma, e_f}$} & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma, e_f})\!:\!S, H, s,
\cred{\clazy\sigma, e_f}$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
  (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \cred{\sigma,
  e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e,
  \cred{\sigma', e_f'}$} &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
  (\psi:\SRETURN~x), \cred{\sigma, e_f}$ \\ $
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma, e_f})\!:\!S, H,  s,
  \cred{\sigma, e_f}$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{Minefield semantics for the language.\label{fig:minefield-semantics}}
\end{center}
\end{figure*}

%% Generated by bibtex from your ~.bib file.  Run latex,
%% then bibtex, then latex twice (to resolve references)
%% to create the ~.bbl file.  Insert that ~.bbl file into
%% the .tex source file and comment out
%% the command \texttt{{\char'134}thebibliography}.
%% % This next section command marks the start of
%% % Appendix B, and does not continue the present hierarchy
%% \section{More Help for the Hardy}
%% The sig-alternate.cls file itself is chock-full of succinct
%% and helpful comments.  If you consider yourself a moderately
%% experienced to expert user of \LaTeX, you may find reading
%% it useful but please remember not to change it.
%% %\balancecolumns % GM June 2007
%% % That's all folks!

%% % {\color{red}
%% %   (Note 3.\ includes $e_\mainpgm$ treated as the body of
%% %   $(\DEFINE\ ({\tt main})\ e_\mainpgm)$.)
%% % }

%% %% =======

%% %------------------------------------------------------------%
%% %\input{lazy-liveness}
}
\end{document}

>>>>

Functional programs use dynamically allocated memory extensively.  The
allocation  is either  explicit (while using  constructors, for example) or
implicit (as while creating  closures).  Programs  written in  lazy
functional languages put additional  demands on memory as they require
closures to  be carried  from the  point of creation  to the  point of
evaluation.

@@

\begin{figure*}[t!]
\begin{center}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
\\
\hline
          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
\rightsquigarrow
$  $\rho', S, H[\rho'(z) := (H(\rho'(x)),H(\rho'(y)))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e$      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
\rightsquigarrow
\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
$\rho', S, H[\rho'(z) := H(\rho'(x)) + H(\rho'(y))], e$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
\sc{funcall} \\
\hline
$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
e)\!:\!S, H,
(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho'(z) := v], e$ &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
\end{figure*}

@@

%------------------------------------------------------------%

@@

\subsubsection{From CFGs to GC-ready DFA}
\label{sec:NFA-approx}
 We use  the algorithm by Mohri  and Nederhof~\cite{mohri00regular} to
 approximate a CFG to a {\em strongly regular\/} grammar.  For
 example, the grammar
 fragment  for  the  non-terminal  $\var{\Df{\length}{1}}$  after  the
 Mohri-Nederhof transformation is:
 \begin{eqnarray*}
   \var{\Df{\length}{1}} &\rightarrow& \clazy\var{\Df{\length}{1'}} \mid
   \acdr\var{\Df{\length}{1}}
   \mid \clazy\var{\Df{\length}{1}}\\
   \var{\Df{\length}{1'}} &\rightarrow& \clazy\var{\Df{\length}{1'}}
   \mid \epsilon
 \end{eqnarray*}

The strongly regular grammar is converted  into a set of NFAs, one for
each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
pairs  of consecutive  edges labelled  \bcar\acar\ or  \bcdr\acdr\ and
constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
simplification is continued till a fixed point is reached, after which
the edges labelled \bcar \ and  \bcdr are deleted.  The details of the
algorithm, its correctness and termination proofs are given by Karkare
et.  al.~\cite{karkare07liveness,asati14lgc}.  The resulting automaton
has  edges labelled  with  \acar,  \acdr\ and  \clazy\  only. In  this
automaton, for every edge labelled \clazy, we check if the source node
of the edge has a valid path to  a final final state.  If yes, we mark
the mark  the source node as  final. Finally, we remove  all the edges
labelled  \clazy\  and  convert  the automaton  into  a  deterministic
automaton.    This   effectively  implements   the   $\hookrightarrow$
simplification rules for  \bcar, \bcdr, and \clazy\  to obtain forward
paths. While  checking for liveness  during GC, a  forward access-path
only if it can reach a final state.
Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
from the grammar for \var{\Df{\length}{1}}.  The final DFA is shown in
Figure~\ref{fig:example-automata}(b).  This expectedly says that given
a  demand $\epsilon$  on  \length,  the liveness  of  its argument  is
$\acdr^{*}$  (the  spine  of   the  list  is  traversed).   Similarly,
Figure~\ref{fig:example-automata}(c)     shows     the     NFA     for
\var{\Lanv{9}{\pa}}.          The         DFA         shown         in
Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
paths.   This  reflects  the  lazy  nature  of  our  language.   Since
\length\ does not  evaluate the member elements of  the argument list,
the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
liveness based GC is triggered beyond $\pi_9$.

@@

\pagebreak
\section{Appendix}

@@

To set up the minefield semantics, we follow these steps:
\begin{enumerate}
\item We enrich  the abstract machine state $\rho,$ $S,$  $ H,$ $e$ to
  $\rho,$ $S,$ $H,$ $e,$ $\sigma,$ $e_f$, where $\sigma$ is the demand
  on the expression $e$, and $e_f$  is the body of the function. $e_f$
  and $\sigma$ are  both needed to calculate  the liveness environment
  for expressions.  The information  in a suspended evaluation context
  is also similarly  augmented with the demand on it  and the function
  body of  which the suspended evaluation  is a part.  Thus  a stacked
  entry now takes the form $(\rho, x, e, \sigma, e_f)$. A modification
  of the small  step semantics which carries the  extra information is
  shown in Figure~\ref{fig:minefield-semantics}.
\item Given  the current  evaluation context $\rho,  S, H,  e, \sigma,
  e_f$ and $LF$, we can construct a liveness environment as follows:
  \begin{enumerate}
  \item If $e$ is an expression, then:  
  $$ \Lv = \bigcup_{\psi \in \mathcal{P}(e)} \mathcal{M}(\psi),
    \text{where }
  \mathcal{M} = \mathcal{L}(e_f, \sigma, \Lfonly)$$
\item If $e$ is an application, then:  
  $$ \Lv = ref(e, \sigma, \Lfonly)$$
  \end{enumerate}
We can define in the same way the liveness environment for each of the
suspended  evaluation contexts in  $S$  giving  a  stack of  liveness
environments that we shall denote $\mathsf{SL}$.
\item $GC$ models a liveness based garbage collection:
$GC(\rho,$ $ S,$ $ H,$ $ L,$ $ SL) = (\rho', S', H')$ 
  \begin{enumerate}
  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
    \notin \Lv$.
\item For each stack entry $(\rho,\_,\_,\_,\_)$ in $S$ with $\Lv'$ as
  the corresponding liveness environment in $\mathsf{SL}$,  and for each $x
  \in domain(\rho)$,  $\rho'(x)=\bot$
  iff $x.\epsilon \notin \Lv'$.
\item For each location $\ell$, $H'(\ell) = \bot$, iff  there is no
  $x$ in either the current environment or one of the stacked
  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
  \ell$  and $x.\alpha \in \Lv$, .
  \end{enumerate}
\end{enumerate}

Starting   from  the   initial  state   $([\;]_\rho,([\;]_\rho,  \ans,
(\print~\ans)):[\;]_{S} ,  [\;]_{H}, (\mainpgm)$, every  transition is
preceded by  a garbage  collection using $GC$.  We consider  a garbage
collection followed  by a transition as  a step. We show  by induction
that, starting  from the  initial state, any  transition of  $n$ steps
occurs without a \bang.

A few observations: First observe that the transitions  which involving dereferencing and can lead to \bang are those which have a $H(\rho(x))$ term in their premise for some variable $x$.

\begin{figure*}[h!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\ 
\hline
\hline
          &\makecell{ $\rho, (\rho', x, e, \cred{\sigma', e_f'})\!:\!S,
  H, \kappa, \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) :=
    \kappa], e, \cred{\sigma', e_f'}$ }   &  \sc{const}
\\
\hline
          & \makecell{$\rho, (\rho', z, e, \cred{\sigma',
    e_f'})\!:\!S, H, (\CONS~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
  \rho', S, H[\rho'(z) := (H(\rho'(x)),H(\rho'(y)))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
  \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \cred{\sigma, e_f}$ \\ $
  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma', e_f'}$}      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
  H, (\CAR~x), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
  (\CAR~x), \cred{\sigma, e_f})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma , \_ }$}      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \makecell{$\rho, (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H,
  (+~x~y), \cred{\sigma, e_f}$ \\  $\rightsquigarrow \rho', S, H[\rho'(z)
    := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma', e_f'}$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
  H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_}$}      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
  S, H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, y,
  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_ }$}      &
\sc{prim-2-clo} \\
\hline
{$\mathit{g}~\mbox{defined as}$
$~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
  (g~\myvec{x}), \cred{\sigma, e_f}$ \\ $\rightsquigarrow [\myvec{y} \mapsto
    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma, e_g}$}      &
\sc{funcall} \\
\hline
$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma, e_f}$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e, \cred{\sigma, e_f}$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma, e_f}$} & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma, e_f}$ \\   $\rightsquigarrow
\rho, S, H,  e_2, \cred{\sigma, e_f}$} & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma, e_f})\!:\!S, H, s,
\cred{\clazy\sigma, e_f}$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
  (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \cred{\sigma,
  e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
  \cred{\sigma', e_f'}$} &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
  (\psi:\SRETURN~x), \cred{\sigma, e_f}$ \\ $
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma, e_f})\!:\!S, H,  s,
  \cred{\sigma, e_f}$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{Minefield semantics for the language.\label{fig:minefield-semantics}}
\end{center}
\end{figure*}
\end{document}
